---
title: "Childhood exposure to non-persistent **endocrine disrupting chemicals** and **multi-omic** profiles: a panel study"
title_textsize: "60px"
main_findings: "Some of these biological signatures point towards the potential biological effects of OP pesticide and phthalate metabolites on the nervous system"
main_fontfamily: "Rasa"
main_textsize: "70px"
main_topsize: 0.05
main_bottomsize: 0.04
logoleft_name: "images/logos_sx.svg"
#logoright_name: "images/logos_dx.svg"
poster_height: "46.8in"
poster_width: "33.1in"
author:
  - name: Lorenzo Fabbri
    affil: a, b
    main: TRUE
    orcid: "0000-0003-3031-322X"
  - name: Ronan Garlantézec
    affil: c
  - name: Karine Audouze
    affil: d
  - name: Mariona Bustamante
    affil: e, a, b, f
  - name: Ángel Carracedo
    affil: g, h
  - name: Leda Chatzi
    affil: j
  - name: Juan Ramón González
    affil: a, j, f
  - name: Regina Gražulevičienė
    affil: k
  - name: Barbara Heude
    affil: l
  - name: Hector Keun
    affil: m
  - name: Chung-Ho E Lau
    affil: n, o
  - name: Eduard Sabidó
    affil: e, b
  - name: Alexandros P Siskos
    affil: m
  - name: Rémy Slama
    affil: p
  - name: Cathrine Thomsen
    affil: q
  - name: John Wright
    affil: r
  - name: Wen Lun Yuan
    affil: l, s
  - name: Maribel Casas
    affil: a, b, f
  - name: Martine Vrijheid
    affil: a, b, f
  - name: Léa Maitre
    affil: a, b, f
affiliation:
  - num: a
    address: Barcelona Institute for Global Health (ISGlobal), Barcelona, Spain
  - num: b
    address: Universitat Pompeu Fabra (UPF), Barcelona, Spain
  - num: c
    address: Univ Rennes, CHU Rennes, Inserm, EHESP, Irset (Institut de recherche en santé environnement et travail), UMR_S 1085, Rennes, France
  - num: d
    address: Université Paris Cité, T3S, INSERM UMR-S 1124, 45 rue des Saints Pères, Paris, France
  - num: e
    address: Center for Genomic Regulation (CRG), Barcelona Institute of Science and Technology (BIST), Barcelona, Spain
  - num: f
    address: CIBER Epidemiologa y Salud Pública (CIBERESP), Madrid, Spain
  - num: g
    address: Medicine Genomics Group, Centro de Investigación Biomédica en Red Enfermedades Raras (CIBERER), University of Santiago de Compostela, CEGEN-PRB3, Santiago de Compostela, Spain
  - num: h
    address: Galician Foundation of Genomic Medicine, Instituto de Investigación Sanitaria de Santiago de Compostela (IDIS), Servicio Gallego de Salud (SERGAS), Santiago de Compostela, Spain
  - num: i
    address: Department of Preventive Medicine, Keck School of Medicine, University of Southern California, Los Angeles, USA
  - num: j
    address: Department of Mathematics, Universitat Autònoma de Barcelona, Bellaterra, Spain,
  - num: k
    address: Department of Environmental Sciences, Vytautas Magnus University, Kaunas, Lithuania
  - num: l
    address: Université de Paris, Centre for Research in Epidemiology and Statistics (CRESS), INSERM, INRAE, Paris, France
  - num: m
    address: Cancer Metabolism & Systems Toxicology Group, Division of Cancer, Department of Surgery and Cancer & Division of Systems Medicine, Department of Metabolism, Digestion & Reproduction, Imperial College London, Hammersmith Hospital Campus, London, UK
  - num: "n"
    address: MRC Centre for Environment and Health, School of Public Health, Imperial College London, London, UK
  - num: o
    address: Division of Systems Medicine, Department of Metabolism, Digestion & Reproduction, Imperial College, South Kensington, London, UK
  - num: p
    address: Team of Environmental Epidemiology applied to Reproduction and Respiratory Health, Institute for Advanced Biosciences (IAB), Inserm, CNRS, Université Grenoble Alpes, Grenoble, France
  - num: q
    address: Department of Environmental Health, Norwegian Institute of Public Health, Oslo, Norway
  - num: r
    address: Bradford Institute for Health Research, Bradford Teaching Hospitals NHS Foundation Trust, Bradford, UK
  - num: s
    address: Singapore Institute for Clinical Sciences (SICS), Agency for Science, Technology, and Research (A*STAR), Singapore, Singapore
author_textsize: "28px"
authorextra_textsize: "26px"
affiliation_textsize: "22px"
column_numbers: 2
body_textsize: "38px"
bibliography: "bibliography.bib"
link-citations: true
csl: "small.csl"
knit:
  pagedown::chrome_print
output:
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

<style>
#main-img-left, #main-img-center, #main-img-right {
 width: 35%;
}
</style>

<style>
.main_bottom {
 background-image: none;
 background-color: #FFFFFF;
}
</style>

<style>
.section h1 {
  font-size: 55px;
}
</style>

<style>
.section h2 {
  font-size: 40px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, 
                      tidy = FALSE, 
                      message = FALSE, warning = FALSE, 
                      fig.align = 'center', out.width = "100%")
```

<br>

# Background & Objectives

* Individuals are exposed to multiple environmental pollutants with endocrine disrupting activity (endocrine disruptors, EDCs)
* Few studies have integrated biomarkers of exposure to EDCs and multi-omics
* We aimed to identify multi-omic signatures associated with childhood exposure to non-persistent EDCs using a network approach

<br>

# Methods

* Data: HELIX (Human Early-Life Exposome) Child Panel Study ($N_A = 140$, $N_B = 143$)

<br>

<img src="images/study_design.svg" 
  style="height: 2000px; width:1500px;"/>

```{r, echo=FALSE}
#knitr::include_graphics(paste0(getwd(), "/images/study_design.svg"))
```

<!-- * We developed Gaussian Graphical Models (GGMs) based on shrinkage estimates of pairwise partial correlations between EDCs and molecular features in each week -->
<!-- * The obtained networks for the two visits were merged in order to identify reproducible associations -->
<!-- * We corroborated the associations between EDCs and omic features with -->
<!--   * Manual literature review (PubMed) -->
<!--   * Adverse health effects associated with the EDCs (PubChem) -->
<!--   * Traits associated with the CpG sites (EWAS Atlas) -->
<!--   * Genes mapped to the CpG sites (EWAS Atlas, EWAS Catalog) -->

<br>

<br>

<br>

# Results

<!-- ## Study population -->

<!-- |**Characteristic**        |   **Visit A**, N = 140   |   **Visit B**, N = 143   | -->
<!-- |:-------------------------|:------------------:|:------------------:| -->
<!-- |__Cohort__                |                    |                    | -->
<!-- |BIB                       |      27 (19%)      |      27 (19%)      | -->
<!-- |EDEN                      |      23 (16%)      |      23 (16%)      | -->
<!-- |KANC                      |      27 (19%)      |      26 (18%)      | -->
<!-- |RHEA                      |      29 (21%)      |      29 (20%)      | -->
<!-- |SAB                       |      34 (24%)      |      38 (27%)      | -->
<!-- |__Sex__                   |                    |                    | -->
<!-- |female                    |      60 (43%)      |      61 (43%)      | -->
<!-- |male                      |      80 (57%)      |      82 (57%)      | -->
<!-- |__Age__                   | 6.97 (6.44, 8.85)  | 7.55 (6.93, 9.62)  | -->
<!-- |__Ethnicity__             |                    |                    | -->
<!-- |Caucasian                 |     129 (92%)      |     132 (92%)      | -->
<!-- |Pakistani                 |     10 (7.1%)      |     10 (7.0%)      | -->
<!-- |Other                     |      1 (0.7%)      |      1 (0.7%)      | -->
<!-- |__zBMI__                  | 0.28 (-0.34, 1.09) | 0.26 (-0.34, 1.09) | -->
<!-- |__Season__                |                    |                    | -->
<!-- |autumn                    |      35 (25%)      |      38 (27%)      | -->
<!-- |spring                    |      50 (36%)      |      49 (34%)      | -->
<!-- |summer                    |      20 (14%)      |      8 (5.6%)      | -->
<!-- |winter                    |      35 (25%)      |      48 (34%)      | -->

## Merged network: reproducible associations across visits

<img src="images/networks_main.svg" 
  style="height: 2000px; width:1500px;"/>

```{r, echo=FALSE}
#knitr::include_graphics(paste0(getwd(), "/images/networks_main.svg"))
```

* The process of network merging led to the exclusion of a significant number of edges, maintaining 950 edges compared to 4,083 and 4,908 in networks A and B, respectively

| Chemical class | Compound | Omic layer | Omic feature | Median pcor | Literature review |
|:---:|:---:|:---:|:---:|:---:|:---:|
| OP pesticides | DEP | Serum metabolome | Serotonin | 0.052 | Exposure to pesticides and changes in the serotonergic system [@slotkin2008developmental; @judge2016mechanism; @sarrouilhe2021exposome] |
| Phenols | TRCS | Serum metabolome | Serotonin | -0.055 | Association between personal care products and serotonin [@houten2016changes] |
| Phthalates | oh-MiNP | Serum metabolome | Kynurenine | -0.056 | Mediating effect of quinolic acid  on exposure to phthalates and neurological disorders [@nassan2019high] |
| Phenols | TRCS | Proteome | Leptin | -0.066 | Effect of mixtures of phenols and parabens [@lee2022race] |

# Conclusions

* We employed an integrative method to investigate reproducible associations between non-persistent EDCs and multi-omic profiles in a child cohort
* Strengths: repeated pooled samples of urines; quantifiable measurements of the metabolome
* Limitations: relatively small sample size; residual confounding
* Some of these biological signatures point towards the potential biological effects of OP pesticides and phthalates on the nervous system

# References
